Theoretical prediction of disrupted Min oscillation in flattened \emph{Escherichia coli}

### 2015-09-24 paper/plos-final.tex
changes for publication

\newcommand{\changed}[1]{{\color{red}#1}}
%\newcommand{\changed}[1]{}

@@

Previous experimental studies have shown that the MinD protein system
is capable of exhibiting oscillations in round
shapes~\cite{corbin2002exploring,fange2006noise} as well as in
connected three pronged tube shapes~\cite{varma2008min}, in which the
oscillations seem to seek out the extreme poles in the cell.
%%   One conclusion of these
%% studies has been that the MinD system robustly forms bipolar
%% oscillations regardless of variations in cellular shape.
M\"annik \emph{et al.} have recently shown that there are limitations to
this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are spatially irregular. Seen with fluorescent microscopy, the
MinD maximize in multiple locations within the cell, in a seemingly
random sequence. These experiments allow for an opportunity to test
models of the MinD system against more extreme cases than have been
seen thus far.  An additional interesting question is whether the
irregularity observed by M\"annik \emph{et al.}  indicates that the MinD
system can become chaotic (as in
e.g. Ref.~\citenum{kuramoto1978diffusion}), or if this irregularity is
the result of a stochastic but nonchaotic process.
%

@@

\begin{figure}
  \includegraphics[width=8.7cm]{reactions}
  \caption{Reactions included in the Min system model of Huang
    \emph{et al.}, which is used in this
    work~\cite{huang2003dynamic}.}\label{fig:reactions}
\end{figure}

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Model and Cell Shapes}
\label{sec:model-method-shapes}
We implement the reaction-diffusion model of Huang \emph{et
  al.}~\cite{huang2003dynamic}.  Figure~\ref{fig:reactions} shows the
reaction process.  The cytoplasmic MinD:ADP complex undergoes
nucleotide exchange and is changed into the MinD:ATP complex.  This
will naturally diffuse and attach to the cell membrane.  A cytoplasmic
MinE will attach to the wall bound MinD:ATP complex and after a time
will activate ATP hydrolysis.  This breaks up the complex, releasing
MinE, phosphate, and MinD:ADP back into the cytoplasm.  The MinD:ADP
will undergo nucleotide exchange and begin again the cyclic process.
The model is defined by a set of five reaction-diffusion equations:

@@

\begin{figure*}
  %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
  \begin{center}
    \hspace{-3.5cm}
    \includegraphics[height=6.4cm]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}\\
    \vspace{-1.5em}
    \hspace{-3.5cm}
    \includegraphics[height=6.4cm]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
    \vspace{-1.5em}
  \end{center}
  \caption{Images of the concentration of each protein species in a
    natural pill-shaped bacterium at one-second time intervals. The
    upper plots shows results from the deterministic model and the
    lower shows results from the stochastic model.  The order of
    frames is such that individual MinD proteins begin at the bottom
    of the plot (in the MinD:ATP state in the cytoplasm), and progress
    upward until they reach the MinE:MinD:ATP membrane-bound complex.
    At that point, they will spontaneously dissociate into cytoplasmic
    MinE (the top row) and the starting state of cytoplasmic
    MinD:ADP.}.
  \label{image-p}
\end{figure*}

@@

\begin{figure}
  \begin{center}
    \includegraphics[height=6.4cm]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/correlation.pdf}
  \end{center}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the pill-shaped cell, shown against
    the correlation time.  Data for both the the deterministic and
    stochastic models are shown.  The stochastic model shows an
    oscillation period of 39.5 seconds and coherence time of 307 seconds.
    The correlation functions are scaled to have the same initial
    value.}
  \label{corr-pill}
\end{figure}

@@

Figure~\ref{image-p} shows a series of color plots of the density of
proteins at each stage of the reaction cycle. Deterministic simulation
data is shown above and stochastic simulation data is shown below.
The cells shown are $4\micron$ in length, measured from end to end.
We have `smeared out' the stochastic concentrations in a manner meant
to reproduce the images shown by diffraction limited fluorescence
microscopy.  We do so using the two dimensional Gaussian approximation
developed by Zhang \emph{et al.}~\cite{zhang2007gaussian}.  In this
approximation we use a numerical aperture value of 1.3, which is the
same as used by M\"annik experimentally, and a wavelength of $650nm$.
Each frame is 2.5 seconds ahead of the last, and each image shows the
concentration of a given state of protein (of the five described in
the reaction model) summed over the coordinate normal to the page.

Figure~\ref{image-p} begins about 300 seconds into the simulation and
shows one period of oscillation.  At $t=0$ there is a high
concentration of MinD:ATP that has accumulated on the membrane at the
bottom of the cell. An important aspect of Huang's model is that the
MinD is attracted to and sticks to the membrane nonlinearly: as it
accumulates there it begins to `recruit' other MinD that is diffusing
in the cytoplasm nearby, causing peaks in concentration to build up on
the walls.  Meanwhile, the MinE creeps downward as it reacts with the
membrane-bound MinD, forming the MinD:MinE:ATP complex, then breaks it
apart and diffuses downward a bit more before it again reacts with
membrane-bound MinD.  This process can be seen in the form of the
well-known ``MinE rings'' (actually, MinE bound to MinD on the
membrane).  These rings are visible in the deterministic model plots
as a green band on the walls from 0 seconds to 4 seconds (and later
from 15 to 23 seconds).  The appearance of these rings in the
deterministic model alongside their less obvious appearance in the
stochastic model highlights an advantage of the deterministic model:
the idealization of deterministic data allows one to see patterns in
the averaged behavior that might otherwise be missed.  In the
stochastic model the MinE still exhibits higher concentrations in the
same regions throughout the process, but what would ideally be a
``ring'' pattern is instead an asymmetric collection of maxima, which
would become a ring after phase-locked averaging.

@@

Both Fig.~\ref{image-p} and its associated movies (see supplementary
material), show that the stochastic model exhibits a ``starry night''
effect characterized by spatially fixed points of protein density
build up, as recruitment leads to clusters of MinD forming on the wall
and then subsequently dissipating.  This result, which is not observed
in experiments, stems from the omission of diffusion along the
membrane from Huang's model, which would allow these recruitment
clusters to spread out.  Experiments have confirmed that diffusion
along the membrane does occur, albeit with a rate two orders of
magnitude lower than that in the cytoplasm~\cite{meacci2006mobility}.
The `stars' in the effect typically last for around 10 seconds. The
membrane-bound proteins will diffuse by a distance of around one
micron during this time, which is enough to spread out across the
polar section of the one micron diameter wild-type pill cells.  As we
see below, the flattened, irregular cells are large enough that this
one micron blurring might not completely eliminate this `starry night'
effect.

@@

We also analyze the temporal periodicity of the system, which we
quantify with the temporal correlation function of the total MinD
found in two opposite polar regions.  This correlation function, which
is displayed in Fig.~\ref{corr-pill}, is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods, indicating that the system exhibits periodic
oscillation, with some stochastic irregularity in the period.
%
In contrast, in the deterministic model the pill-shaped cell exhibits
complete coherence, indicating that its behavior is perfectly
periodic.

\begin{figure*}
  \centering
  \hspace{-3cm}
  \vspace{1cm}
  \includegraphics[width=16cm]{../paper/plot-ave}
  \caption{Arrows depicting successive maxima in space
    and time overlayed on a color plot of the total MinD density
    averaged over the same time period.  The simulation time covered
    for the flattened cells is 350 seconds, which is the same
    period of time depicted in the experimental data plots of M\"annik
    \emph{et al.}~\cite{mannik2012robustness}.  For the wild-type pill
    shape, we only cover 250 seconds, in order to provide a
    useful comparison due to its shorter oscillation period.  The top
    row shows plots published by M\"annik \emph{et al.} of the MinD maxima behavior
    and the bottom two rows show our simulations using the stochastic
    and deterministic models, respectively.  We simulated
    approximations to the two shapes observed by M\"annik, which we call
    \emph{shape A} and \emph{shape B}.  In addition, we studied two
    flattened stadium shapes which we call \emph{stadium A} and
    \emph{stadium B} corresponding in aspect ratio and thickness to
    the two experimental shapes.  The spatial length scale of all the
    figures shown was identical. Each of the flattened cell shapes
    uses the same color scale for the number of proteins per unit
    area.  Finally, we display the natural pill shape, which was also
    featured in Figs.~\ref{image-p} and~\ref{corr-pill}, with a
    different color scale to reflect the thicker cell containing more
    proteins per cross-sectional area.  }
  \label{randst-plot-ave}
\end{figure*}

%% \begin{figure}
%%   \begin{center}
%%     \includegraphics[height=6.4cm]{../data/shape-randst/0_40-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}\\
%%     \includegraphics[height=6.4cm]{../data/shape-stad/0_40-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
%%   \end{center}
%%   \caption{Temporal correlation function of the total MinD found in
%%     two opposite polar regions of the shape \emph{A} (above) and the
%%     stadium \emph{A} (below) cell shapes, shown against the
%%     correlation time.  Data for both the the determinisitic and
%%     stochastic models are shown, and the correlation functions are
%%     scaled to have the same initial value.  For shape \emph{A}, the
%%     stochastic model shows an oscillation period of 52 seconds and
%%     coherence time of 288 seconds, so that it takes roughly 5.5
%%     periods for the behavior to decohere.  For the stadium \emph{A},
%%     the same model shows an oscillation period of 48.4 seconds and
%%     coherence time of 638 seconds, so that it takes roughly 13
%%     periods for the behavior to decohere.}
%%   \label{fig:corr-pancake-A}
%% \end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=8.7cm]{../data/shape-randst/0_40-18_60-28_60-94_00-15_00-full_array/plots/correlation.pdf}
    \includegraphics[width=8.7cm]{../data/shape-stad/0_40-2_92-1_18-0_00-15_00-full_array/plots/correlation.pdf}
  \end{center}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the shape \emph{B} (above) and the
    stadium \emph{B} (below) cell shapes, shown against the
    correlation time.  Data for both the the deterministic and
    stochastic models are shown, and the correlation functions are
    scaled to have the same initial value.  For shape \emph{B}, the
    stochastic model shows an oscillation period of 58 seconds and
    coherence time of 306 seconds, so that it takes roughly 5.3 periods
    for the behavior to decohere.  For the stadium \emph{B}, the same
    model shows an oscillation period of 51 seconds and coherence time
    of 442 seconds, so that it takes roughly 8.6 periods for the
    behavior to decohere.}
  \label{fig:corr-B}
\end{figure}

\subsection*{Comparison of Experimental and Stadium Shapes}
%\label{sec:model-method-shapes}
As explained above, we focus on simulation of four illustrative
flattened cell shapes: two shapes created to replicate those shapes
observed experimentally by M\"annik \emph{et
  al.}~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two `stadium' shapes that have the
identical aspect ratio, thickness and volume (stadium \emph{A} and
stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.  All of the
flattened cells have lost the rotational symmetry of the original pill
shapes, but while shape~\emph{A} and shape~\emph{B} have only one
mirror plane symmetry (that is normal to the flattened plane),
stadium~\emph{A} and stadium~\emph{B} have three.  This allows us to
distinguish between the effect of flattening the cell and the cell
shape's irregularity and asymmetry.

@@

From each of these flattened simulations, we have chosen a typical 350
second segment to compare with the published results of
M\"annik \emph{et al.}~\cite{mannik2012robustness}, exemplified in arrow plots in
which the arrow heads show the location of sequential MinD maxima
within the cell (in Fig.~\ref{randst-plot-ave}).  In addition to
arrows between successive maxima in space and time, we plot as a
colored background the density of MinD proteins averaged over the same
time period.  We note that we have manually verified that our
(computer-generated) arrow plots also reflect a human interpretation
of a movie of the same data.  Finally, for comparison we present the
same plot for a wild-type cell, with a time period of 250 seconds to
account for its short period of oscillation.

@@

The predictions of the stochastic model show some similarity to the
experimental results in terms of irregularity of the locations of
maxima, but the stochastic model displays significantly less
variation in the direction of oscillations than experiment, suggesting
inadequacies in the model.  We also note that the stadium shapes in
Fig.~\ref{randst-plot-ave} appear to have maxima location irregularity
that is qualitatively similar to our irregular shapes: in both
stochastic simulations the overall oscillation is largely bipolar, but
with irregularity in the locations of the maxima that form at each
end.

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
the flattening and accompanying lateral expansion of the cell versus
cell shape irregularity and asymmetry in the temporal regularity of
the oscillations.  Although in the movies the stadiums appear to
display, on average over long time scales, a somewhat more regular
oscillation than appears in the irregular shapes \emph{A} and
\emph{B}, the limited time range of the arrow plots (350 seconds) is
inadequate to make this comparison.  We therefore return to the
correlation function between the number of proteins in a segment at
the top of the cell and the number of proteins in a segment at the
bottom of the cell.  The ``top'' and ``bottom'' regions for computing
the correlation time were established by dividing the long axis into
thirds.  This resulted in regions at the top and bottom ranging from
19\% to 33\% of the cell volume for shapes \emph{A} and \emph{B}, with
the stadium shapes featuring 30\% of the volume in the top and bottom
regions.  Fig.~\ref{fig:corr-B} shows this plot for shape \emph{B} and
stadium \emph{B}.  Here we plot correlations of correlation times up
to 500 seconds, a biologically relevant timescale.  \changed{\emph{E. coli}} cell
doubling times vary widely depending on environmental conditions, but
will often be between 20 minutes to 100
minutes~\cite{pierucci1972chromosome}, while fluorescent microscopy
experiments have shown that the FtsZ ring builds itself with a half
life of roughly 30 seconds~\cite{stricker2002rapid}.

@@

\bibliography{paper-MinD}

>>>>

%\newcommand{\changed}[1]{{\color{red}#1}}
\newcommand{\changed}[1]{#1}

@@

Previous experimental studies have shown that the MinD protein system
is capable of exhibiting oscillations in round
shapes~\cite{corbin2002exploring,fange2006noise} as well as in
connected three pronged tube shapes~\cite{varma2008min}, in which the
oscillations seem to seek out the extreme poles in the cell.
%%   One conclusion of these
%% studies has been that the MinD system robustly forms bipolar
%% oscillations regardless of variations in cellular shape.
M\"annik \emph{et al.} have recently shown that there are limitations to
this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are spatially irregular. Seen with fluorescent microscopy, the
MinD maximize in multiple locations within the cell, in a seemingly
random sequence. These experiments allow for an opportunity to test
models of the MinD system against more extreme cases than have been
seen thus far.  An additional interesting question is whether the
irregularity observed by M\"annik \emph{et al.}  indicates that the MinD
system can become chaotic (as in
e.g. Ref.~\cite{kuramoto1978diffusion}), or if this irregularity is
the result of a stochastic but nonchaotic process.
%

@@

\begin{figure}
  %% \includegraphics[width=8.7cm]{reactions}
  \caption{Reactions included in the Min system model of Huang
    \emph{et al.}, which is used in this
    work~\cite{huang2003dynamic}.}\label{fig:reactions}
\end{figure}

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Model and Cell Shapes}
\label{sec:model-method-shapes}
We implement the reaction-diffusion model of Huang \emph{et
  al.}~\cite{huang2003dynamic}.  Fig~\ref{fig:reactions} shows the
reaction process.  The cytoplasmic MinD:ADP complex undergoes
nucleotide exchange and is changed into the MinD:ATP complex.  This
will naturally diffuse and attach to the cell membrane.  A cytoplasmic
MinE will attach to the wall bound MinD:ATP complex and after a time
will activate ATP hydrolysis.  This breaks up the complex, releasing
MinE, phosphate, and MinD:ADP back into the cytoplasm.  The MinD:ADP
will undergo nucleotide exchange and begin again the cyclic process.
The model is defined by a set of five reaction-diffusion equations:

@@

\begin{figure*}
  \begin{center}
    %% \hspace{-3.5cm}
    %% \includegraphics[height=6.4cm]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}\\
    %% \vspace{-1.5em}
    %% \hspace{-3.5cm}
    %% \includegraphics[height=6.4cm]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
    %% \vspace{-1.5em}
  \end{center}
  \caption{Images of the concentration of each protein species in a
    natural pill-shaped bacterium at one-second time intervals. The
    upper plots shows results from the deterministic model and the
    lower shows results from the stochastic model.  The order of
    frames is such that individual MinD proteins begin at the bottom
    of the plot (in the MinD:ATP state in the cytoplasm), and progress
    upward until they reach the MinE:MinD:ATP membrane-bound complex.
    At that point, they will spontaneously dissociate into cytoplasmic
    MinE (the top row) and the starting state of cytoplasmic
    MinD:ADP.}.
  \label{image-p}
\end{figure*}

@@

\begin{figure}
  \begin{center}
    %% \includegraphics[height=6.4cm]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/correlation.pdf}
  \end{center}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the pill-shaped cell, shown against
    the correlation time.  Data for both the the deterministic and
    stochastic models are shown.  The stochastic model shows an
    oscillation period of 39.5 seconds and coherence time of 307 seconds.
    The correlation functions are scaled to have the same initial
    value.}
  \label{corr-pill}
\end{figure}

@@

Fig~\ref{image-p} shows a series of color plots of the density of
proteins at each stage of the reaction cycle. Deterministic simulation
data is shown above and stochastic simulation data is shown below.
The cells shown are $4\micron$ in length, measured from end to end.
We have `smeared out' the stochastic concentrations in a manner meant
to reproduce the images shown by diffraction limited fluorescence
microscopy.  We do so using the two dimensional Gaussian approximation
developed by Zhang \emph{et al.}~\cite{zhang2007gaussian}.  In this
approximation we use a numerical aperture value of 1.3, which is the
same as used by M\"annik experimentally, and a wavelength of $650nm$.
Each frame is 2.5 seconds ahead of the last, and each image shows the
concentration of a given state of protein (of the five described in
the reaction model) summed over the coordinate normal to the page.

Fig~\ref{image-p} begins about 300 seconds into the simulation and
shows one period of oscillation.  At $t=0$ there is a high
concentration of MinD:ATP that has accumulated on the membrane at the
bottom of the cell. An important aspect of Huang's model is that the
MinD is attracted to and sticks to the membrane nonlinearly: as it
accumulates there it begins to `recruit' other MinD that is diffusing
in the cytoplasm nearby, causing peaks in concentration to build up on
the walls.  Meanwhile, the MinE creeps downward as it reacts with the
membrane-bound MinD, forming the MinD:MinE:ATP complex, then breaks it
apart and diffuses downward a bit more before it again reacts with
membrane-bound MinD.  This process can be seen in the form of the
well-known ``MinE rings'' (actually, MinE bound to MinD on the
membrane).  These rings are visible in the deterministic model plots
as a green band on the walls from 0 seconds to 4 seconds (and later
from 15 to 23 seconds).  The appearance of these rings in the
deterministic model alongside their less obvious appearance in the
stochastic model highlights an advantage of the deterministic model:
the idealization of deterministic data allows one to see patterns in
the averaged behavior that might otherwise be missed.  In the
stochastic model the MinE still exhibits higher concentrations in the
same regions throughout the process, but what would ideally be a
``ring'' pattern is instead an asymmetric collection of maxima, which
would become a ring after phase-locked averaging.

@@

Both Fig~\ref{image-p} and its associated movies (see supplementary
material), show that the stochastic model exhibits a ``starry night''
effect characterized by spatially fixed points of protein density
build up, as recruitment leads to clusters of MinD forming on the wall
and then subsequently dissipating.  This result, which is not observed
in experiments, stems from the omission of diffusion along the
membrane from Huang's model, which would allow these recruitment
clusters to spread out.  Experiments have confirmed that diffusion
along the membrane does occur, albeit with a rate two orders of
magnitude lower than that in the cytoplasm~\cite{meacci2006mobility}.
The `stars' in the effect typically last for around 10 seconds. The
membrane-bound proteins will diffuse by a distance of around one
micron during this time, which is enough to spread out across the
polar section of the one micron diameter wild-type pill cells.  As we
see below, the flattened, irregular cells are large enough that this
one micron blurring might not completely eliminate this `starry night'
effect.

@@

We also analyze the temporal periodicity of the system, which we
quantify with the temporal correlation function of the total MinD
found in two opposite polar regions.  This correlation function, which
is displayed in Fig~\ref{corr-pill}, is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods, indicating that the system exhibits periodic
oscillation, with some stochastic irregularity in the period.
%
In contrast, in the deterministic model the pill-shaped cell exhibits
complete coherence, indicating that its behavior is perfectly
periodic.

\begin{figure*}
  \centering
  %% \hspace{-3cm}
  %% \vspace{1cm}
  %% \includegraphics[width=16cm]{../paper/plot-ave}
  \caption{Arrows depicting successive maxima in space
    and time overlayed on a color plot of the total MinD density
    averaged over the same time period.  The simulation time covered
    for the flattened cells is 350 seconds, which is the same
    period of time depicted in the experimental data plots of M\"annik
    \emph{et al.}~\cite{mannik2012robustness}.  For the wild-type pill
    shape, we only cover 250 seconds, in order to provide a
    useful comparison due to its shorter oscillation period.  The top
    row shows plots published by M\"annik \emph{et al.} of the MinD maxima behavior
    and the bottom two rows show our simulations using the stochastic
    and deterministic models, respectively.  We simulated
    approximations to the two shapes observed by M\"annik, which we call
    \emph{shape A} and \emph{shape B}.  In addition, we studied two
    flattened stadium shapes which we call \emph{stadium A} and
    \emph{stadium B} corresponding in aspect ratio and thickness to
    the two experimental shapes.  The spatial length scale of all the
    figures shown was identical. Each of the flattened cell shapes
    uses the same color scale for the number of proteins per unit
    area.  Finally, we display the natural pill shape, which was also
    featured in Figs~\ref{image-p} and~\ref{corr-pill}, with a
    different color scale to reflect the thicker cell containing more
    proteins per cross-sectional area.  }
  \label{randst-plot-ave}
\end{figure*}

\begin{figure}
  \begin{center}
    %% \includegraphics[width=8.7cm]{../data/shape-randst/0_40-18_60-28_60-94_00-15_00-full_array/plots/correlation.pdf}
    %% \includegraphics[width=8.7cm]{../data/shape-stad/0_40-2_92-1_18-0_00-15_00-full_array/plots/correlation.pdf}
  \end{center}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the shape \emph{B} (above) and the
    stadium \emph{B} (below) cell shapes, shown against the
    correlation time.  Data for both the the deterministic and
    stochastic models are shown, and the correlation functions are
    scaled to have the same initial value.  For shape \emph{B}, the
    stochastic model shows an oscillation period of 58 seconds and
    coherence time of 306 seconds, so that it takes roughly 5.3 periods
    for the behavior to decohere.  For the stadium \emph{B}, the same
    model shows an oscillation period of 51 seconds and coherence time
    of 442 seconds, so that it takes roughly 8.6 periods for the
    behavior to decohere.}
  \label{fig:corr-B}
\end{figure}

\subsection*{Comparison of Experimental and Stadium Shapes}
%\label{sec:model-method-shapes}
As explained above, we focus on simulation of four illustrative
flattened cell shapes: two shapes created to replicate those shapes
observed experimentally by M\"annik \emph{et
  al.}~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two `stadium' shapes that have the
identical aspect ratio, thickness and volume (stadium \emph{A} and
stadium \emph{B}), shown in Fig~\ref{randst-plot-ave}.  All of the
flattened cells have lost the rotational symmetry of the original pill
shapes, but while shape~\emph{A} and shape~\emph{B} have only one
mirror plane symmetry (that is normal to the flattened plane),
stadium~\emph{A} and stadium~\emph{B} have three.  This allows us to
distinguish between the effect of flattening the cell and the cell
shape's irregularity and asymmetry.

@@

From each of these flattened simulations, we have chosen a typical 350
second segment to compare with the published results of
M\"annik \emph{et al.}~\cite{mannik2012robustness}, exemplified in arrow plots in
which the arrow heads show the location of sequential MinD maxima
within the cell (in Fig~\ref{randst-plot-ave}).  In addition to
arrows between successive maxima in space and time, we plot as a
colored background the density of MinD proteins averaged over the same
time period.  We note that we have manually verified that our
(computer-generated) arrow plots also reflect a human interpretation
of a movie of the same data.  Finally, for comparison we present the
same plot for a wild-type cell, with a time period of 250 seconds to
account for its short period of oscillation.

@@

The predictions of the stochastic model show some similarity to the
experimental results in terms of irregularity of the locations of
maxima, but the stochastic model displays significantly less
variation in the direction of oscillations than experiment, suggesting
inadequacies in the model.  We also note that the stadium shapes in
Fig~\ref{randst-plot-ave} appear to have maxima location irregularity
that is qualitatively similar to our irregular shapes: in both
stochastic simulations the overall oscillation is largely bipolar, but
with irregularity in the locations of the maxima that form at each
end.

However Fig~\ref{randst-plot-ave} leaves unclear the importance of
the flattening and accompanying lateral expansion of the cell versus
cell shape irregularity and asymmetry in the temporal regularity of
the oscillations.  Although in the movies the stadiums appear to
display, on average over long time scales, a somewhat more regular
oscillation than appears in the irregular shapes \emph{A} and
\emph{B}, the limited time range of the arrow plots (350 seconds) is
inadequate to make this comparison.  We therefore return to the
correlation function between the number of proteins in a segment at
the top of the cell and the number of proteins in a segment at the
bottom of the cell.  The ``top'' and ``bottom'' regions for computing
the correlation time were established by dividing the long axis into
thirds.  This resulted in regions at the top and bottom ranging from
19\% to 33\% of the cell volume for shapes \emph{A} and \emph{B}, with
the stadium shapes featuring 30\% of the volume in the top and bottom
regions.  Fig~\ref{fig:corr-B} shows this plot for shape \emph{B} and
stadium \emph{B}.  Here we plot correlations of correlation times up
to 500 seconds, a biologically relevant timescale.  \changed{\emph{E. coli}} cell
doubling times vary widely depending on environmental conditions, but
will often be between 20 minutes to 100
minutes~\cite{pierucci1972chromosome}, while fluorescent microscopy
experiments have shown that the FtsZ ring builds itself with a half
life of roughly 30 seconds~\cite{stricker2002rapid}.

@@

%% \bibliography{paper-MinD}
\begin{thebibliography}{10}
\providecommand{\url}[1]{\texttt{#1}}
\providecommand{\urlprefix}{URL }
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi
\providecommand{\bibAnnoteFile}[1]{%
  \IfFileExists{#1}{\begin{quotation}\noindent\textsc{Key:} #1\\
  \textsc{Annotation:}\ \input{#1}\end{quotation}}{}}
\providecommand{\bibAnnote}[2]{%
  \begin{quotation}\noindent\textsc{Key:} #1\\
  \textsc{Annotation:}\ #2\end{quotation}}
\providecommand{\eprint}[2][]{\url{#2}}

\bibitem{adams2009bacterial}
Adams DW, Errington J (2009) Bacterial cell division: assembly, maintenance and
  disassembly of the z ring.
\newblock Nature Reviews Microbiology 7: 642--653.
\bibAnnoteFile{adams2009bacterial}

\bibitem{lutkenhaus2007assembly}
Lutkenhaus J (2007) Assembly dynamics of the bacterial mincde system and
  spatial regulation of the z ring.
\newblock Annu Rev Biochem 76: 539--562.
\bibAnnoteFile{lutkenhaus2007assembly}

\bibitem{shen2010examination}
Shen B, Lutkenhaus J (2010) Examination of the interaction between ftsz and
  mincn in e. coli suggests how minc disrupts z rings.
\newblock Molecular microbiology 75: 1285--1298.
\bibAnnoteFile{shen2010examination}

\bibitem{shapiro2009and}
Shapiro L, McAdams HH, Losick R (2009) Why and how bacteria localize proteins.
\newblock Science 326: 1225--1228.
\bibAnnoteFile{shapiro2009and}

\bibitem{yu1999ftsz}
Yu XC, Margolin W (1999) Ftsz ring clusters in min and partition mutants: role
  of both the min system and the nucleoid in regulating ftsz ring localization.
\newblock Molecular microbiology 32: 315--326.
\bibAnnoteFile{yu1999ftsz}

\bibitem{meacci2005min}
Meacci G, Kruse K (2005) Min-oscillations in escherichia coli induced by
  interactions of membrane-bound proteins.
\newblock Physical biology 2: 89.
\bibAnnoteFile{meacci2005min}

\bibitem{huang2003dynamic}
Huang KC, Meir Y, Wingreen NS (2003) {Dynamic structures in Escherichia coli:
  spontaneous formation of MinE rings and MinD polar zones}.
\newblock Proceedings of the National Academy of Sciences 100: 12724--12728.
\bibAnnoteFile{huang2003dynamic}

\bibitem{kerr2006division}
Kerr RA, Levine H, Sejnowski TJ, Rappel WJ (2006) Division accuracy in a
  stochastic model of min oscillations in escherichia coli.
\newblock Proceedings of the National Academy of Sciences of the United States
  of America 103: 347--352.
\bibAnnoteFile{kerr2006division}

\bibitem{mannik2009bacterial}
M{\"a}nnik J, Driessen R, Galajda P, Keymer JE, Dekker C (2009) Bacterial
  growth and motility in sub-micron constrictions.
\newblock Proceedings of the National Academy of Sciences 106: 14861--14866.
\bibAnnoteFile{mannik2009bacterial}

\bibitem{corbin2002exploring}
Corbin BD, Yu XC, Margolin W (2002) Exploring intracellular space: function of
  the min system in round-shaped escherichia coli.
\newblock The EMBO journal 21: 1998--2008.
\bibAnnoteFile{corbin2002exploring}

\bibitem{fange2006noise}
Fange D, Elf J (2006) Noise-induced min phenotypes in e. coli.
\newblock PLoS computational biology 2: e80.
\bibAnnoteFile{fange2006noise}

\bibitem{varma2008min}
Varma A, Huang KC, Young KD (2008) The min system as a general cell geometry
  detection mechanism: Branch lengths in y-shaped escherichia coli cells affect
  min oscillation patterns and division dynamics.
\newblock Journal of bacteriology 190: 2106--2117.
\bibAnnoteFile{varma2008min}

\bibitem{mannik2012robustness}
M{\"a}nnik J, Wu F, Hol FJ, Bisicchia P, Sherratt DJ, et~al. (2012) Robustness
  and accuracy of cell division in escherichia coli in diverse cell shapes.
\newblock Proceedings of the National Academy of Sciences 109: 6957--6962.
\bibAnnoteFile{mannik2012robustness}

\bibitem{mannik2010bacteria}
M{\"a}nnik J, Sekhavati F, Keymer J, Dekker C (2010) Bacteria in submicron
  channels and microvalves.
\newblock Proc of MicroTAS 2010, Groningen, The Netherlands : 1346--1348.
\bibAnnoteFile{mannik2010bacteria}

\bibitem{kuramoto1978diffusion}
Kuramoto Y (1978) Diffusion-induced chaos in reaction systems.
\newblock Progress of Theoretical Physics Supplement 64: 346--367.
\bibAnnoteFile{kuramoto1978diffusion}

\bibitem{meinhardt2001pattern}
Meinhardt H, de~Boer PA (2001) Pattern formation in escherichia coli: a model
  for the pole-to-pole oscillations of min proteins and the localization of the
  division site.
\newblock Proceedings of the National Academy of Sciences 98: 14202--14207.
\bibAnnoteFile{meinhardt2001pattern}

\bibitem{kruse2002dynamic}
Kruse K (2002) A dynamic model for determining the middle of< i> escherichia
  coli</i>.
\newblock Biophysical journal 82: 618--627.
\bibAnnoteFile{kruse2002dynamic}
